---
title: 'Measures of network position and structure '
author: Shane A. Scaggs
date: '2022-09-30'
slug: measures-of-network-position-and-structure
categories:
  - methods
tags:
  - structure
  - nodes
  - matrix
  - analysis
draft: yes
---

Many networks that we find interesting are large and structurally complex. This makes them difficult to visually understand. And although coming up with ways to visualize such networks is a kind of fun, most of the time we need practical ways to describe the position and structure of complex networks without dealing with the tangled hairball. 

This post covers some a variety of ways to measure network position and structure. By position, I mean properties at the node level, such as how many connections a node has (degree centrality) or how often it lies on the shortest path (betweenness). By structure, I mean properties of subgraphs that are found within the network as whole, such as triads, communities, or spectra. 

# Prerequistes 

To start, we need to generate some networks to work with. Let's use the three canonical network models: the Erdos-Renyi random graph, the Watts-Strogatz small world, and the Albert-Barbasi preferential attachment network. We can simulate these using `igraph`.

```{r}
library(igraph)
```

Let each network have 500 nodes. 

```{r}
n = 500
```

Each model has different parameters. The random graph is simplest -- we define the number of nodes and the probability that they are connected. 

```{r}
rg = erdos.renyi.game(n=n, p=0.02)
rg
```

The small world model has more parameters. To make our graphs somewhat comparable, we will keep the size equal to n. The `nei` parameter controls how many neighbors a node is connected to. The parameter `p` controls the probability that the a connection is rewired. When `p` is 0, the graph forms a circlular chain. 

```{r}
sw = watts.strogatz.game(dim=1, size=n, nei = 5, p=0.2)
sw
```

Finally the prefernential attachment network has 

```{r}
barabasi.game(n=n, power=1, m=5)
```


