---
title: '(Re)Introduction to Networks in R '
author: Harrison Fried, Shane A. Scaggs
date: '2022-10-7'
slug: re-introduction-to-networks-in-r
categories:
  - methods
tags: 
toc: yes
draft: yes
---

In this meeting, we cover the basics of working with networks in R. By the end of the workshop, you will be acquainted with three packages -- `igraph`, `statnet`, and `tidygraph` -- and you will be able to create a network from scratch, turn it into a network object, create vertex and edge attributes, and genereate a basic visualization. Let's get started. 

# R Markdown Introduction

Before we begin talking about networks, let's first talk about the R studio environment. Most R code can be successfully run using classic `.r` files.

Sometimes, it makes more sense to present information and structure R code using R Markdown `.rmd`, which is what this type of file is.

R Markdown contains the same capabilities as classic `.r` files, but has the added ability to insert text (and equations!) that can be knit into a clean-appearing .html or .pdf file for presentations or write-ups. You also have the ability to hide or show chunks of code in the "knit" file. Therefore, R markdown provides a nice format for embedding R output within a research paper (such as by only including R plots within text). Or, R markdown is great to write tutorials or show coding logic, by showing R code within text.

It's easy to add headers (use #'s).

# Example Header A

## Example Subheader B

### Example Tertiary Header C

Then, it's easy to embed R code "chunks", which always follow this basic format:

```{r}
a <- 1+1
a
```

Now that we've discussed R markdown basics, let's talk about network data.

# Data Structures 

Before we dive into network objects, let's understand the data structures that are used to represent relational data. We will focus on **matrices** and **edgelists**, as these are among the most common structures used for network data. 

# Adjacency matrix
## Adjacency matrix: using toy data

```{r, echo=F }
v <- c("i", "j", "k")
v
M <- matrix(c(0,0,0,
              0,0,0,
              0,0,0), 
            nrow = length(v), 
            dimnames = list(v,v))
M
```

One way to represent a network is using a square matrix where every node in the network has its own row and column. To do this, let's create a vector of three nodes `i`, `j`, and `k`. 

Every cell in the matrix `M` represents a relationship between the two nodes -- these are the network edges. A value of `0` means that no edge is present. So in this matrix, there aren't any edges. But let's add an edge from `i` to `j` and from `i` to `k`.

```{r}
M[1,2] <- 1
M[1,3] <- 1
M
```

This is the conventional way of representing **directed** edges between nodes. We can represent a reciprocal edge from `k` to `i` as well. In an **undirected** network, this information is redundant. 

```{r}
M[3,1] <- 1
M
```

Cells along the diagonal are known as loops. For example, we can set an edge from `j` to `j` along the diagonal.

```{r}
M[2,2] <- 1
M["j","j"] <- 1 #Same
M
```

You can always extract the diagonal using `diag()`. This is also a handy way to set the diagonal to `0` if your network does not contain loops. 

```{r}
diag(M) <- 0
class(M)
M
```

## Adjacency matrix: using simulated data (bernoulli)

In our example matrix above there were `3` nodes and a matrix with `9` possible edges. These means that for $N$ nodes, there are $N^2$ possible edges. We can use this information to simulate a random network. 

```{r}
set.seed(777)
# How many nodes? 
N <- 7 

# How many edges? 
N_edges <- N^2

# Use binomial distribution to simulate edges (coin flip)
simM <- matrix(rbinom(N_edges, 
                   size=1,
                   prob=0.5),
            nrow = N)
# No loops
diag(simM) <- 0
simM
```

## Adjacency matrix: using real data

Let's take a look at a sample data-set of 19 environmental issues, and their connections to each other - let's check what this looks like in excel.

First, let's set our working directory so that R knows where to find our file, and where to save output to. 
```{r}
setwd("~/Desktop/SENG") #Set working directory'
issuenet <- read.csv("IssueIssue_Data.csv", header=T)
class(issuenet)
```

Now, let's inspect the data as it is, and ensure it's a matrix.
```{r}
dim(issuenet) #19 rows, 19 columns! A square.
class(issuenet)
issuenet <- as.matrix(issuenet) 
head(issuenet)
tail(issuenet)
```

There aren't row names, currently. Let's fix that.
```{r}
rownames(issuenet) <- colnames(issuenet) #Add row names
head(issuenet)
```

Let's now add an "i_" to the beginning of each of our issue nodes. This will help us differentiate issue nodes from other node types later on.
```{r}
colnames(issuenet) <- rownames(issuenet) <- paste0("i_", colnames(issuenet))
head(issuenet)
```

Now let's make sure that the diagonal values aren't NA.
```{r}
diag(issuenet) <- 0
issuenet
#View(issuenet) #Can't run View() in R markdown knit
```


#Edgelist
## The Edgelist: Using toy data

An **edgelist** is a more compact way to store network data. This is because it does not include the `0`s in that are shown in the matrix. Instead an edgelist only contains rows for each of the edges that are present in the network. This may seem trivial for a small network of 3 nodes, but when you work with a large network with hundreds or thousands of rows, eliminating all of those zeros is very convenient. 

To construct an edgelist, stack rows on top of one another using the `rbind` function. 
```{r}
E <- rbind(c('i','k'),
           c('j','j'),
           c('k','i'))
class(E)
colnames(E) <- c('Sender','Receiver')
E
class(E)
```

You can think of the first column as a *sender* column. The second column is a *receiver* column. This is how directionality is represented in the edge list. 

If you want to add edges to an edgelist, simply bind a new row to the current edgelist. 

```{r}
E <- rbind(E, c('i','j'))
E
```

Use the same bracket `[]` notation to index, such as here, where we delete the fourth row. 

```{r}
E <- E[-4,] 
E
```

## But Why? 

You might be wondering... why do we need to know these details about data structure? Here are two reasons. 

First, whether you collect your own network data or receive data from someone else, you'll need to store/wrangle the data into one of these formats. When you collect your own data, you can store it in this format right away. But if you receive a network you did not collect, knowing these data structures will help you understand what you are looking at. 

Second, understand data structure opens the door to simulation. Put differently, you don't need to collect data to begin learning about and visualizing networks. You can simulate synthetic data. Here is a brief example. 

## Edgelist: Using simulated data (bernouli graph simulation)

We can use a similar procedure to simulate an edgelist. We just need to create a third column to simulate the edges, then filter out the zeros. 
```{r}
set.seed(777)
N <- letters[1:5]
#or
N <- c("a","b","c","d","e")
simE <- expand.grid(N,N)
# ?expand.grid #help on this function.
simE
simE$Edge <- rbinom(nrow(simE), 
                 size = 1, 
                 prob = 0.5)
dim(simE)
head(simE) #notice 3 columns, including presence/absence of edge
```

Now, let's make this a true Edgelist, by removing rows where there is not an edge (i.e., value == 0)
```{r}
simE <- simE[!simE$Edge == 0 & !simE$Var1 == simE$Var2, 1:2] #Only include edges that have a value, and that aren't loops, only show Columns 1 and 2
head(simE)
```

## Edgelist: using real data

Let's load in an edgelist - but first, let's check out what it looks like in Excel.
```{r}
aa <- read.csv("actor-actor_edgelist.csv")
head(aa)
aa$send <- paste0("a_", aa$send)
aa$receive <- paste0("a_", aa$receive)
```

Let's inspect the data, then fix node names, which will help with future analyses.
```{r}
class(aa)
aa <- as.matrix(aa)
dim(aa)
head(aa)
```


# Network objects in `igraph`

Now we will turn these data structures into `igraph` network objects. First, install and load the `igraph` package.

```{r, echo=FALSE,message=F, warning=F}
#install.packages('igraph')
library(igraph)
```

First let's create some attributes for our `3` node network. They don't need to be fancy; just some sizes and colors. 

```{r}
att <- data.frame(
    name = c("i","j","k"), 
    size = c(20,27,34),
    color = c('tomato', 
              'cornflowerblue', 
              'darkorchid')
)
att
```

## Matrix conversion to network

To create a network object using a matrix, we use the `graph.adjacency` function. In graph theory, relational matrices are often referred to as adjacency matrices. Use the `M` matrix that we created before.

```{r}
M
class(M)
gM <- graph.adjacency(M)
class(gM)
gM
```
When we call the object, we receive a summary of the network. `DN` means this is a *directed network* with `3` nodes and `3` edges. There is just one attributes, the `name` of the vertices. We can tell that this is a categorical vertex attribute by the notation `(v/c)`. The edges are listed at the bottom. 

Now, let's manually add vertex attributes to the network `gM` from out `att` data frame. 
```{r}
gM <- set_vertex_attr(gM, name = 'size', value = att$size)
gM <- set_vertex_attr(gM, name = 'color', value = att$color)
gM
```

## Edgelist conversion to network

To create an object using an edgelist, we use the `graph.data.frame` function. Using edgelists is nice because you can add the vertex attribute table in directly. 

```{r}
E
class(E)
gE <- graph.data.frame(E)

#Add in the attributes
gE <- set_vertex_attr(gM, name = 'size', value = att$size)
gE <- set_vertex_attr(gM, name = 'color', value = att$color)

#Or, with one line of code to add attributes while we make igraph object:
gE <- graph.data.frame(E, vertices = att)
gE
class(gE)
```

Once you have an `igraph` network object, you can extract vertex and edge attributes. 
```{r}
V(gE)$color
V(gE)$size
```
We don't have any edge attributes in this network, but if we did, we would use `E(object)$attr` syntax. 

## Visualize `igraph` objects 

There are many aesthetic properties of network graph. But you can generally think of them as **vertex** (or node) aesthetics, **edge** aesthetics, and **network** aesthetics. To learn more about all of the ways to specify these in `igraph`, called `??igraph.plotting`. 

Let's visualize our objects `gE` and `gM`. Here is a generic plot. 

```{r}
plot(gE)
plot(gM)
```

By default, `igraph` labels each node with the `$name` attribute, and, if they exist, it will choose the sizes and colors using the `$color` and `$size` attributes. Since we set `size` and `color` in our `att` data frame, we replaced the defaults. But we can easily override them. 

```{r}
plot(gE, 
     vertex.color = '#2a00fa', 
     vertex.size = 30)
```

Let's also change the color of the label to make it easier to read, change the edge colors to black, and make the arrow heads a bit smaller. 

```{r}
plot(gE, 
     vertex.color = '#2a00fa',
     vertex.size = 30,
     vertex.label.color = 'white', 
     edge.color = 'black',
     edge.arrow.size = 0.681)
```

Finally, let's set the layout to `layout.circle` and add some curve to the edges.

```{r}
plot(gE, 
     vertex.color = '#2a00fa',
     vertex.size = 30,
     vertex.label.color = 'white', 
     edge.color = 'black',
     edge.arrow.size = 0.681, 
     edge.curved = 0.1,
     layout = layout.circle)
```


## Network objects in `statnet`

Create network objects in `statnet` is pretty intuitive. First, install and load the `statnet` suite of packages. 

```{r, eval=FALSE,echo=F, message=F, warning=F}
#install.packages('statnet', dependencies=T)
library(statnet)
```

To create network objects, use the `network` function. When using an edgelist, tell the function you are using one in the `matrix.type` parameter.
```{r}
library(statnet)
M
E
att
netM <- network(M, vertex.attr = att)
netE <- network(E, vertex.attr = att, 
                matrix.type = "edgelist" )
netM
netE
c(class(netM), class(netE))
```

Like the `igraph` object, when we call the `network` object, we get a summary of some important network details. 

To extract variables from the `network` object, we use one of three operators:

- `%v%` -- extract vertex attributes
- `%e%` -- extract edge attributes 
- `%n%` -- extract network attributes 

For example, we can see the values for color by extracting the attribute from the network by calling the network object, the appropriate operator, and the name of the attribute. 

```{r}
netE %v% 'color'
```


## Convert real data to `igraph` and `network` objects
### igraph conversion

```{r}
aa_igraph <- graph.data.frame(aa)
class(aa_igraph)
plot(aa_igraph)
dim(aa)
```
Let's plot the full network.

```{r}
plot(aa_igraph)
#messy!
```

Let's change some plot arguments to improve our graph.
```{r}
plot(aa_igraph, 
     vertex.color = '#2a00fa',
     vertex.size = 3,
     vertex.label = "",
     edge.color = 'black',
     edge.arrow.size = 0.02, 
     edge.curved = 0.1)
```

Or, we can just look at a subset of the network.

```{r}
aa_shortened <- aa[1:100,]
dim(aa_shortened)
aa_igraph_shortened <- graph.data.frame(aa_shortened)
```

Let's plot this new, shortened network using igraph.
```{r}
plot(aa_igraph_shortened, 
     vertex.color = '#2a00fa',
     vertex.size = 7,
     vertex.label = "",
     edge.color = 'black',
     edge.arrow.size = 0.02, 
     edge.curved = 0.1)
```

What about plotting our issue network?
```{r}
issuenet
class(issuenet)
gissuenet <- graph.adjacency(issuenet, weighted = T)
plot(gissuenet, 
     vertex.color = '#2a00fa',
     vertex.size = 7,
     vertex.label = "",
     edge.color = 'black',
     edge.arrow.size = 0.02, 
     edge.curved = 0.1, 
     edge.width=edge.betweenness(gissuenet)/10)
```

```{r}
par(mfrow=c(2,2)) 
plota <- plot(gissuenet, 
     vertex.color = '#2a00fa',
     vertex.size = 7,
     vertex.label = "",
     edge.color = 'black',
     edge.arrow.size = 0.02, 
     edge.curved = 0.1, 
     edge.width=edge.betweenness(gissuenet)/10,
     layout=layout.circle,
     main="circle")
plotb <- plot(gissuenet, 
     vertex.color = '#2a00fa',
     vertex.size = 7,
     vertex.label = "",
     edge.color = 'black',
     edge.arrow.size = 0.02, 
     edge.curved = 0.1, 
     edge.width=edge.betweenness(gissuenet)/10,
     layout=layout.sphere,
     main="sphere")
plotc <- plot(gissuenet, 
     vertex.color = '#2a00fa',
     vertex.size = 7,
     vertex.label = "",
     edge.color = 'black',
     edge.arrow.size = 0.02, 
     edge.curved = 0.1, 
     edge.width=edge.betweenness(gissuenet)/10,
     layout=layout.random,
     main="random")
plotd <- plot(gissuenet, 
     vertex.color = '#2a00fa',
     vertex.size = 7,
     vertex.label = "",
     edge.color = 'black',
     edge.arrow.size = 0.02, 
     edge.curved = 0.1, 
     edge.width=edge.betweenness(gissuenet)/10,
     layout=layout.fruchterman.reingold,
     main="fruchterman.reingold")
```
