---
title: A Brief Introduction to R 
author: "Harrison Fried"
date: '2021-04-05'
slug: a-brief-introduction-to-r
categories:
  - methods
tags:
  - Rstudio
  - coding
  - analysis
  - workflow
---


```{r setup, include=F, echo=F}
knitr::opts_chunk$set(echo = TRUE)
```


## Module 1: R Studio Basics

R is both a language and computing environment.
R is open source and free for anyone to download and use.

### Panels

There are four visible panels at a time, each with different purposes
(1) The R console accepts commands and produces output to your commands.
(2) A text editor window to create and record  code.
(3) An environment window that shows your stored data in R
(4) A plot/help/files window that can alternate to show your plots, help reauests, and file directory

For example, use the text editing window to write the following code:

```{r Plot the point 2,2}
plot(2,2)
```
In R Studio, notice that the code is stored in the text editor (script), is passed to the console, and then is plotted in the plot window
Also, panels can be resized by dragging the slider

### Getting Help

Stuck on a line of code? Not sure what a function does exactly? 
Want an example of someone else using that function?
Use the `?` before any function to pull up its help page in the `Help` window.

```{r Get help on the function "Plot"}
?plot
```

### Executing Commands

Send your code from the text editor/script to the console in one of three ways:
(1) Code > Run Selected Line(s)
(2) Select Run Icon on top-right of Script Editor box
(3) Keyboard Shortcut 
Mac: Command + Return
Windows: Ctrl + Enter



## Module 2: More R basics


### Scripts

You can have multiple scripts in R
Scripts appear as tabs in the text edit panel
Save your scripts OFTEN
R studio will sometimes crash unexpectedly and you could lose unsaved changes

### Annotating

Have you wondered why some of the text in your script is green colored? 
It's to annotate your script!
Annotations are saved in the text editor window
You can use them to section your script and to write notes about what your code is doing.
You SHOULD annotate your script, so you can go back potentially YEARS later and still understand what you did (why did I write that code?)

Anything after the HASHTAG becomes annotated, for example:

```{r Annotation_e}
2+2 #An easy math problem 
2 #Math    + 2
```

Notice in the second math line above, the "+2" is not added because it occurs after the #

### Types of Objects

R data is stored as Objects

Common types of R Objects:
Vectors: Single Dimension (a variable or column)
Matrices: a data matrix
Data Frames: Basically tables
Data frames are COMMON and are organized like an excel worksheet.
-Variables are columns
-Observations are rows

```{r mtcars_df}
mtcars
?mtcars #mtcars is a pre-loaded dataset always available to anyone in R.
class(mtcars) #Class function is useful to show the class of the object (what type of object?)
```

Other (LESS COMMON) object types:
Multiple Dimension - Can think of as tabs in excel file
-Lists
Other 2 or 3-D object classes include
-Arrays
-Tables


### Creating your own objects

Assign a name to your object using `<-`, `->` or `=`
Most people just use `<-`

```{r Creating_objects}
a <- 1 + 2 + 3
a

5 + 5 -> b
b

c = log(5)
c
```

You can set an R object to equal another object, or as a function of another object

```{r Manipulating_objects}
x = a
x
y = b * 10
y
```

The function `c()` stands for concatenate and is used to create a vector
Notice you can name your object almost anything

```{r Naming_objects}
vec <- c(1,2,3,4,5,6)
vec
```

Naming objects IS case sensitive

Useful arithmetic functions in Base R, which we can use for `vec`
```{r Arithmetic}
mean(vec)
```
Also, consider using: `min()`; `max()`; `summary()`

### Working with a Data Set

Let's return to `mtcars` dataset
The object of the dataset is called `mtcars`
```{r mtcars2}
mtcars
class(mtcars) # Class of dataset -- data frame
str(mtcars) # Structure of the data - useful to examine the data set
head(mtcars) ## Return first 6 lines of data frame
tail(mtcars) ## Return last 6 lines of data frame
length(mtcars) #Essentially produces the number of columns
length(mtcars$mpg) #Essentially produces the number of rows
dim(mtcars) #The order used in R is always: Rows, Columns (Left to right!)
names(mtcars) ## Return names of all variables
```
In the above code, we ran multiple lines of code to help us understand the structure of the dataset, each with a unique function.

### Vectors

We've already discussed using the concatenate function `c()` to make a vector

```{r vector1}
new_vec <- c(1,3,5,7)
length(new_vec)
```

Now let's make a vector of actual data, which we can pull from the `mtcars` dataset

```{r mtcars_vector}
miles_per_gallon <- mtcars$mpg #The '$' refers to variables (columns) within the dataset
miles_per_gallon
```

### Basic Indexing

Ever wonder what the square brackets do in R? `[]`
Wonder no more.
The square brackets are used to index, which means to pull out subsections from objects.
```{r Indexing_vectors}
practice_vec <- c(0,10,20,30,40,50,60,70)
practice_vec
indexed_vec <- practice_vec[1:4] #Observations 1 thru 4
indexed_vec
indexed_vec <- practice_vec[5] #Only the 5th observation
indexed_vec
indexed_vec <- practice_vec[-5] #Everything but the 5th observation
indexed_vec
```

Notice that the object `indexed_vec` becomes updated every time we run a new line.
Indexing a vector is straightforward, we will return to indexing for 2-D data frames.

## Data Frames

`mpg` is another base R dataset.

```{r mpg_dataset}
#install.packages("ggplot2")
library(ggplot2) #We will return to installing and loading packages later in this document.
mpg #It's not quite a data frame yet
?mpg #What is this data set?
mpg <- data.frame(mpg) #Coerce to a data frame
```

When you index a data frames, remember to use a comma!
When indexing, the first value is for ROWS, and the second is for COLUMNS.

Return the 10th row in 9th variable 'hwy' 
```{r Index_Data_Frame}
mpg[10,9] #Rows, Columns
mpg$hwy[10] #Same as above!
```

Return all rows of the 2nd variable.
```{r ex2}
mpg[,2]
```

Return all columns of the 2nd row
```{r ex3}
mpg[2,]
```

Return all columns in the first 4 rows
```{r ex4}
mpg[1:4,]
```

Return all observations in the 1st, 3rd and 5th variable
```{r ex5}
mpg[, c(1,5,9)]
```

Return all columns when `model` equals `camry`
```{r ex6}
mpg[mpg$model=="camry",]
```
The above line reads: "in the dataframe `mpg`, return all columns for the rows in which the model type is a camry."

There are two important lessons from the above line
Since `model` is a factor, it needs quotes.
Second, we need to use a double equal, '=='. 
R reads `=` as assigning a name to an object. R reads `==` as a logical function, essentially saying, "if it satisfies this condition."

Return all variables with city mpg above 28
```{r ex7}
mpg[mpg$cty>30, ]
```

Return all variables with city mpg greater than or equal to 28
```{r ex8}
mpg[mpg$cty >=28, ]
```

Create a new dataframe object of your indexing!
```{r ex9}
x <-  mpg[mpg$cty>30, ]
x
```

### Load data (Set working directory)

Every time you use R studio, (beginning of a new session), you need to set the working directory
This tells R where on your computer to import/ export data to/from.

There are two ways to set your R working directory:
(1) The GUI menu
(2) Use the command `setwd()`

My working directory: `setwd("/Users/harrisonfried/Box Sync/Midwest CCA coupled networks/ActorForumAnalysis/Data")`
To see your current working directory, use:
```{r directory}
getwd()
```
Now R knows where to get my data from and save what I produce!
Pull a data set in from your working directory using `read.csv()`

### Install and load packages

Packages are a vital feature of R
Most are created by people in your research community!
Packages are modules with one or more functions (that were custom made by the package developers)
Packages are what makes R powerful.

To use a package, you need to first INSTALL it, then LOAD it.
You only have to install packages ONCE (it saves the package into your R studio memory)
You must load them every time you restart R (so R knows what package(s) you will be drawing from today)

Simply install packages with the command:
```{r pckg}
#install.packages("statnet")
```

A common error is that people forget to put the name of the package in quotation marks.
`statnet` is a macro-package for network analysis - it includes many smaller packages within it that each have a purpose in network analysis.

Now, we load the package because it is installed using the `library()` function
```{r lib}
library(statnet)
```


## Module 3: A Primer to Advanced Techniques 

### Creating Functions

You can make your own functions to carry out specific tasks. 
```{r func_ex1}
function(x) { 
    #do something 
}
```

Here is an example.
Suppose you have an equation and you want to give it different values of X to learn the output Y.
This happens when you want to make predictions from a statistical model. 

Load the `iris` dataset, available to anyone with base R.
```{r Iris}
data("iris")    
head(iris)
unique(iris$Species)
```

Let's plot the relationship between petal length and width 

```{r petal}
plot(iris$Petal.Width, iris$Petal.Length, col=iris$Species)
```

Next, let's fit a linear model predicting petal length from width and species using `lm()`
```{r linear}
fit1 <- lm(Petal.Length ~ Petal.Width + Species,  data = iris)
summary(fit1)
```

So what is the equation for this model? 
```{r Equation1}
# PL = a + b1*Petal.Width + b2*versicolor + b3*virginica 
```

Let's see the intercept and slopes 
```{r Intercept}
coefs <- coef(fit1)
coefs
```

Now, let's create a function that embeds the coefficients into a linear equation
```{r LinearEQ}
fit1_predict <- function(PW, versicolor, virginica) { 
    Y <- coefs[1] + coefs[2]*PW + coefs[3]*versicolor + coefs[4]*virginica
    print(Y)
}
```

Test the function with random values for each variable!
```{r Function_te}
fit1_predict(PW=4, versicolor = 0, virginica = 1)
```

Compute predictions for different species  
```{r Predictions}
preds <- fit1_predict(PW=seq(0,3,by=0.2), versicolor = 0, virginica = 0)   # setosa
preds1 <- fit1_predict(PW=seq(0,3,by=0.2), versicolor = 1, virginica = 0)  # versicolor
preds2 <- fit1_predict(PW=seq(0,3,by=0.2), versicolor = 0, virginica = 1)  # virginica 

# Start with a blank canvas 
plot(NULL, xlim=c(0,3), ylim=c(1,7), 
     xlab="Petal.Width", ylab="Petal.Length")

# Add observed values 
points(Petal.Length ~ Petal.Width, data = iris, col=Species)

# Add predictions 
points(seq(0,3,by=0.2), preds, col=1, pch=19)
points(seq(0,3,by=0.2), preds1, col=2, pch=19)
points(seq(0,3,by=0.2), preds2, col=3, pch=19)
```



### for Loops and Apply Statements

A `for` loop will perform actions following a sequence or index, for example 

```{r Forloop}
x <- 1:10
x
for(i in x) { 
    x2 <- i^2
    print(x2)
}
```

For example, for loops can be helpful when working with multiple networks 
```{r Forloops_net}
library(igraph)

# simulate network 

par(mfrow=c(1,1)) #Set the plotting dimensions
g <- barabasi.game(n=40, power = 1, directed = F)
plot(g, vertex.label=NA)

# First, let's use a for loop to create 4 networks with different powers 
pow <- seq(0, 1, length.out=4)
pow

# create a container for the networks 
L <- list()

# loop through different values of pow and plot them
set.seed(1)
par(mfrow=c(2,2), mar=c(1,1,1,1))
for(i in seq_along(pow)) {
    
    # simulate preferential attachment
    L[[i]] <- barabasi.game(n=40, power = i, directed = F)
    
    # plot each network
    plot(L[[i]], 
         edge.arrow.size=0.2, 
         vertex.label=NA)
}
```

Now we can run some descriptive statistic on the list `L` 
One way to do it is to use a loop. 

```{r Loopstats}
for(i in seq_along(L)) { 
    dens <- graph.density(L[[i]])
    apl <- average.path.length(L[[i]])
    print(c(dens, apl))
}
```
Sometimes, the same goal can be accomplished more easily by using an apply statement (`lapply`)

Apply statements take the following form: `lapply(list, function)`

```{r lapply}
lapply(L, graph.density)

# returns a list, so unlist() and put in data frame  
DF <- data.frame(
    Density = unlist(lapply(L, graph.density)),
    APL = unlist(lapply(L, average.path.length))
)
DF
```

`lapply()` is used for lists 
There are also matrix and data.frame versions 

For example, `apply(array, margin, ...)`
Or, `apply(DF, 2, mean)` # this will apply the function `mean()` to every column 
