---
title: 'Autumn 2022 Schedule '
author: 'SENG'
date: '2022-08-27'
slug: autumn-2022-schedule
categories:
  - general
tags:
  - informational
---

## Meeting Information 

- **Time**: 3PM Eastern Standard Time
- **Dates**: 9/9, 9/23, 10/7, 10/21, 11/4, 11/18, 12/2 (see below)
- **Zoom**: [Join meeting](https://osu.zoom.us/j/7886600344?pwd=ZmYweUt6MFBvRlYvdHhjeU9CZnhlZz09)
- **Passcode**: 002244
- **Location**: TBD 

Beginning in September and ending in August, SENG is holding seven hybrid meetings. 

The focus of our meetings this term is on building confidence in our network analysis methods. We will do this by workshopping a lot and hosting fewer seminar style talks. We want our members to end this term feeling even better about using and teaching network analysis in the future. 

After a welcome meeting on 9/9, we will meet to discuss *research methods and design*, focusing on methods used by social scientists to collect complete and personal networks. We will follow this meeting with a refresher on how to work with *networks in* `R`. The next two meetings will involve a deep dive into *Exponential Random Graph Models* (ERGMs). In these meetings we will use simulation, workshop analyses that are members are currently developing, and learn how to interpret and summarize multiple models. The final two meetings will delve into some advanced topics including *Bayesian inference* and *community detection*. 

Date                                                 Topic              Location
------------------------  --------------------------------  --------------------
2022-09-09                                         Welcome!           Zoom & TBD
2022-09-23                                 Research methods           Zoom & TBD
2022-10-07                                    Networks in R           Zoom & TBD
2022-10-21                                    ERGMs, Part I           Zoom & TBD
2022-11-04                                   ERGMs, Part II           Zoom & TBD
2022-11-18                               Bayesian inference           Zoom & TBD
2022-12-02                              Community detection           Zoom & TBD          


```{r, echo=F, message=F, warning=F, include=F}
md <- data.frame(
    Date = as.Date(c('2022-09-16',
                     '2022-09-23',
                     '2022-10-14',
                     '2022-10-28',
                     '2022-11-18',
                     '2022-12-09')), 
    Topic = c('Network research methods and design', 
              'Network analysis in R',
              'Exponential Random Graph Models, Part I',
              'Exponential Random Graph Models, Part II',
              'Bayesian network inference',
              'Community detection'), 
    Location = paste('Zoom', sep = ' & ', 'TBD')
)

library(kableExtra)
library(knitr)
md %>%
    kable('html') %>%
    kable_styling(position = 'center') %>%
    column_spec(1, width = '1cm') 
```

The table above provides the dates, topics, and locations for each of these meetings this autumn. Below you'll find a calendar that shows every Friday when we plan to hold a SENG meeting. For the most part, our meetings fall on the 1st and 3rd Friday each month. 


```{r, echo=F, message=F, warning=F,fig.height=9, fig.width=9, fig.align='left'}
library(tidyverse)

library(showtext)

#font_add_google(
#  name = "Atkinson Hyperlegible", 
#  family = "atkinson"
#)
#
#showtext_auto()



# Creating our data frame 
d <- data.frame(date=seq(as.Date("2022-08-01"), as.Date("2022-12-31"), 1))
d$day <- factor(strftime(d$date, format = '%a'), levels = rev(c('Mon','Tue','Wed','Thu','Fri','Sat','Sun')))
d$week <- factor(strftime(d$date, format = '%V'))
d$month <- factor(strftime(d$date, format = '%B'), levels = c('August','September','October','November','December'))
d$ddate <- factor(strftime(d$date, format = '%d'))

# seng dates 
seng_dates <- as.Date(c('2022-09-09',
                        '2022-09-23',
                        '2022-10-07',
                        '2022-10-21',
                        '2022-11-04',
                        '2022-11-18',
                        '2022-12-02'))
seng_topic <- c('Welcome!',
                'Research methods',
                'Network in R',
                'ERGMs I',
                'ERGMs II',
                'Bayesian networks',
                'Community detection')



# add holidays 
holidates <- as.Date(c('2022-09-05',
                       '2022-11-11',
                       '2022-11-24',
                       '2022-11-25',
                       '2022-12-23',
                       '2022-12-26'))
holiday <- c('Labor day',
             'Indigenous peoples day (observed)',
             'Thanksgiving break',
             'Thanksgiving break',
             'Holiday break',
             'Holiday break')

d$event <- factor(ifelse(d$date %in% seng_dates, 1, 
                         ifelse(d$date %in% holidates, 2, 
                                ifelse( d$day == 'Sun' | d$day == 'Sat', 3, 4))))

d$topic <- ""
d[ d$event == 1, 'topic'] <- seng_topic
d[ d$event == 2, 'topic'] <- holiday


#library(ggrepel)
showtext_opts(dpi=250)




d %>%
    #filter(!day == 'Sat') %>%
    #filter(!day == 'Sun') %>%
    filter(!month == 'August') %>%
    ggplot(aes(x=day, y=week)) + 
    geom_tile(aes(fill=event), color=NA) + 
    geom_text(aes(label=ddate, color=event), show.legend = F) + 
    #geom_text_repel(aes(label=topic)) + 
    scale_fill_manual(values = c('#3300ff','#AD003f','#00c994','#00000011'), labels = c('SENG','Holiday','Weekend','None')) + 
    scale_color_manual(values = c('white','white','black', '#00000088'), labels = c('SENG','Holiday','Weekend','None')) + 
    facet_wrap(~month, scales = 'free', ncol=2) + 
    theme(legend.position = 'top',
          panel.grid = element_blank(), 
          panel.border = element_blank(), 
          plot.background = element_blank(), 
          panel.background = element_rect(color='white',fill='white'),
          axis.text = element_text(color='black', face = 'bold'), 
          strip.background = element_rect(color='black', fill='black'),
          strip.text = element_text(color = 'white', size=14), 
          axis.ticks = element_blank(), 
          text = element_text(size = 16)) + 
    labs(y=NULL, x=NULL, color=NULL, fill=NULL) +
    scale_y_discrete(labels = NULL) + 
    coord_flip()

```

