---
title: Plants, Pollinators, and Projections
author: Shane A. Scaggs
date: '2022-08-01'
slug: plants-pollinators-and-projections
categories:
  - methods
tags:
  - analysis
  - igraph
  - structure
  - bipartite
  - ecological network
bibliography: ppp.bib
draft: yes
---




# Preamble

A bipartite network is a special kind of network. It contains two distinct sets of nodes, often labeled $U$ and $V$, with edges that can only occur *between* $U$ and $V$. Bipartite networks are relatively common in network analysis because there are many interactions that naturally fit the bipartite graph structure. 

The example we'll focus on today is about plant-pollinator interactions. I am going to show you how to manage bipartite network data, visualize it, create bipartite projections, and make inferences about the structure of the bipartite network. Along the way we will answer some ecological questions about plants and pollinators. 

You'll need these R packages to follow along: 

```{r, message=F, warning=F}
library(tidyverse)  # data wrangling 
library(igraph)     # network objects 
library(tidygraph)  # network wrangling
library(ggraph)     # visualization 
```

```{r, echo=F, message=F, warning=F}
source("~/Shane's Projects/visualization tools/color palettes.R")
tallnwhite = theme(panel.background = element_rect(fill='white',color='black', size=1.6),
          plot.background = element_rect(fill=NA, color=NA), 
          strip.background = element_rect(fill='white', color='black', size=1.6),
          strip.text.x = element_text(margin = margin(0.6,0,0.6,0, "cm")),
          panel.grid = element_blank())
```
# Imagine

For this example, imagine you are an ecologist with a desire to understand the world of plants and pollinators. You have a few pollination questions that you are particularly keen to answer: 

1. Which plants host the greatest (and least) number of pollinators? 
2. Do pollinators compete over some plant species? 
3. Which plant-pollinator relationships are vulnerable to extinction if one or the other is lost?

To answer these questions, you've decided to observe plants and painstakingly document which pollinators visit the plants over the course of your study period. With these raw data in hand, you can now represent plant-pollinator interactions as a bipartite network.  

# Data 

The data I'm using for this example was retrieved from the [Web of Life](https://www.web-of-life.es/) database, and was published in a 2002 paper by Dicks, Corbet, and Pywell [-@dicks2002compartmentalization]. You can download the data yourself using the Web of Life map tool by locating the data set in the United Kingdom and dowloading it as a `csv`. 

Like other networks, this one is stored in a matrix. For bipartite networks, the rows represent one set of nodes, and the columns represent the other, making this a rectangular matrix rather than a square one. 

```{r, echo=F, warning=F, message=F}
d0 = read.csv('data/M_PL_062.csv')
```

This matrix is pretty big -- there are `r dim(d0)[2]` insects interacting with `r dim(d0)[1]` plants -- and it needs a bit of wrangling. First thing I'm going to do is `gather` this matrix into an edgelist and rename the columns. 

```{r}
d1 = d0 %>% 
    gather(key=key, value=value, -X) %>%
    select(Plant=X, Insect=key, Observed=value) 
head(d1)
```
Now I want to filter out all of the rows where `Observed == 0`. This will greatly reduce the size of the data.frame, which currently has `r nrow(d1)` rows.  

```{r}
d2 = d1 %>% filter(!Observed == 0) %>% select(Plant, Insect)
```

Before I make this edgelist into an igraph object, I want to create a small tibble to store vertex attributes. Right now the only attributes I have are the species name and whether it is a plant or insect, which are analogous to the bipartite node *types*. Types are required for bipartite matrices and are coded as a Boolean `TRUE` or `FALSE`. 

```{r}
plants = unique(d2$Plant)
insects = unique(d2$Insect)

v = tibble(
    name = c(plants, insects), 
    type = ifelse(name %in% plants, TRUE, FALSE)
)
```

Now we can make the bipartite graph. 

```{r}
b = graph.data.frame(d2, directed = F, vertices = v)
b
```

Because I've already included the required `type` attribute, igraph uses the undirected bipartite network label, `UN-B`. If you don't already have a type attribute in a vertex table, you can add one after you make the network, like this: 

```{r}
b = graph.data.frame(d2, directed = F)
V(b)$type = ifelse( V(b)$name %in% d2[,1], TRUE, FALSE)
b
```

# Visualizing and describing the plant-pollinator network

Next, let's build up a decent visualization of the plant-pollinator network. This network has `r ecount(b)` edges. This will create some visualization challenges, but we'll see what we can do. First, let's plot a basic graph of the igraph object `b` and see how bad it is. 

```{r, include=F, eval=F}
b %>% 
    as_tbl_graph() %>%
    ggraph() + 
    geom_edge_link() + 
    geom_node_point()
```

Whew, this is quite a hairball. In this situation, I find that the `"sugiyama"` layout provides better results, especially when the transparency of the edges is also increased (i.e, `alpha = 0.1`). I'm also going to color the nodes by their `type` and calculate their degree centrality so that I can map the node size onto degree centrality in the visualization. 

```{r, include=F, eval=F}
b %>% 
    as_tbl_graph() %>%
    activate(nodes) %>%
    mutate(degree = centrality_degree()) %>%
    ggraph('sugiyama') + 
    geom_edge_link(alpha=0.1) + 
    geom_node_point(aes(color=type, size=degree), show.legend = F) + 
    theme(panel.background = element_rect(fill='white')) + 
    scale_size_continuous(range = c(0.5,4)) + 
    scale_color_manual(values = c('#3300ff',"#00F99F"))
```
Despite it's imperfections, this layout suggests that there are definitely some species that are more generalist than others. Let's look at this more explicitly by plotting the degree distributions. 

```{r, echo=F, message=F, warning=F}
vd = b %>% 
    as_tbl_graph() %>%
    activate(nodes) %>%
    mutate(degree = centrality_degree(), 
           role = ifelse(type == T, 'plant', 'pollinator')) %>%
    data.frame() 


above100 = vd[ vd$degree > 100, 'name' ]


vd %>% 
    filter(degree > 0) %>%
    ggplot() + tallnwhite + 
    geom_histogram(aes(x=degree, color=role, fill=role), 
                   show.legend = F, alpha=0.5, binwidth = 1) + 
    facet_wrap(~role, scales = 'free_x') + 
    scale_fill_manual(values = c('#006644',"#3300ff")) + 
    scale_color_manual(values = c('#006644',"#3300ff"))
    
```

The number of insects that a plant is pollinated by ranges from around 20 to almost 60, with plants typically playing host to around 30 insects. The number of plants an insect pollinates has a heavy tail: 0.6% of the insects pollinate a combined `r length(unique(d2[ d2$Insect %in% above100, 'Plant']))` plants, which is nearly 40\% of the dataset. The insect that is the most general -- a true fly (*Spallanzania hesperidarum*) -- pollinates 157 plants, which is `r round(157/1044, 2)`\% of all plants in the dataset. 

# Projections 

Each insect in this network has a specific set of direct and indirect edges to which it is connected. By looking at all of the direct connections, we can describe the extent of niche overlap between any two species.  

Likewise, plants may share some pollinators in common. Shared pollinator partners manifest as triangle within the network, and these triangles provide information about which plants are vulnerable to the loss of specific pollinators. 

We can retrieve this information by doing to bipartite projects. 

```{r}
bp = bipartite.projection(b)
p_insect = bp$proj1
p_plants = bp$proj2
```

A projection shows which species are connected to one another through a node in common. For example. if two bees pollinate the same plant, we can represent this affiliation with an edge between the two bees. Likewise, if two plants flower at the same time of year and are subsequently pollinated by a particular bee, we could draw an edge between the two plants who share the bee in common. Two plants/pollinators who share several of the same partners will have an edge between them that is more heavily weighted. 

```{r}
p_insect %>%
    as_tbl_graph() %>%
    ggraph() + tallnwhite + 
    geom_edge_link(aes(x=weight)) + 
    geom_node_point()


p_plants %>%
    as_tbl_graph() %>%
    ggraph() + tallnwhite + 
    geom_edge_link(aes(x=weight)) + 
    geom_node_point()
```


# References 


